---
title: "Finding Similarities between Texts"
engine: knitr
format:
  html:
    fig-width: 10
    fig-height: 12
    dpi: 300
editor_options: 
  chunk_output_type: inline
---

In our previous episodes, we explored word frequencies, lexical density, readability metrics, and TF-IDF to understand individual characteristics of texts. Now we turn to a fundamental question in computational text analysis: **How similar are different texts to each other?**

Text similarity is crucial for many applications: detecting plagiarism, finding texts by similar authors, grouping documents by topic, or discovering literary influences. In this episode, we'll explore two key mathematical approaches for measuring textual similarity: **cosine distance** and **Euclidean distance**.

::: {.callout-note title="Why Similarity Matters"}
Measuring text similarity allows us to:
- **Compare authorship styles** across different works
- **Group texts by genre or theme** based on vocabulary patterns
- **Identify literary influences** between authors and time periods
- **Detect document clustering** for large corpus analysis
- **Explore stylistic evolution** within an author's work over time
:::

## Understanding Distance and Similarity

When we measure text similarity, we're actually measuring how "close" or "distant" texts are in a mathematical space. Each text can be represented as a vector (a list of numbers), where each dimension represents a word and its value represents how important that word is in the text.

For this analysis, we'll use **TF-IDF vectors** - representing each text by the TF-IDF scores of all words in our corpus. This gives us a nuanced view that considers both word frequency and distinctiveness.

### The Mathematics of Text Distance

**Distance** and **similarity** are inversely related:
- **Small distance** = **High similarity** (texts are alike)
- **Large distance** = **Low similarity** (texts are different)

We'll explore two fundamental distance metrics:

1. **Cosine Distance**: Measures the angle between text vectors (focuses on proportional word usage)
2. **Euclidean Distance**: Measures the straight-line distance between text vectors (considers both proportions and magnitudes)

## Setting Up Our Analysis

```{r}
#| echo: false
#| output: false

library(tidyverse)
library(tidytext)
library(gutenbergr)
library(textstem)
library(ggdendro)
library(cluster)
library(corrplot)
```

Let's work with our established corpus, focusing on a subset for clearer visualization. We'll use texts from different genres to see how mathematical similarity aligns with literary categories:

```{r}
# Select a strategically chosen subset for similarity analysis
# Based on the cosine similarity matrix, these texts show more variation and interesting patterns
similarity_corpus <- tribble(
  ~author, ~title, ~year, ~glID, ~genre,
  # Social Fiction - but with some internal variation
  "Jane Austen", "Pride and Prejudice", 1813, 1342, "Social Fiction",
  "Charles Dickens", "Great Expectations", 1861, 1400, "Social Fiction", 
  "Mark Twain", "Tom Sawyer", 1876, 74, "Adventure", # Shows interesting patterns
  # Gothic - distinctive vocabulary
  "Mary Wollstonecraft Shelley", "Frankenstein; Or, The Modern Prometheus", 1818, 84, "Gothic",
  "Bram Stoker", "Dracula", 1897, 521, "Gothic",
  "Charlotte Brontë", "Jane Eyre", 1847, 145, "Gothic",
  # Fantasy - should cluster together
  "Lewis Carroll", "Alice's Adventures in Wonderland", 1865, 11, "Fantasy",
  "Kenneth Grahame", "The Wind in the Willows", 1908, 27805, "Fantasy",
  # Science Fiction - technical vocabulary
  "Jules Verne", "Twenty Thousand Leagues Under the Seas", 1870, 2488, "Science Fiction",
  "H. G. Wells", "The Time Machine", 1895, 43, "Science Fiction",
  # Domestic Fiction - should show similarity to social fiction
  "Elizabeth Gaskell", "Cranford", 1853, 394, "Domestic Fiction",
  "Louisa May Alcott", "Little Women", 1868, 37106, "Domestic Fiction"
)

# Download and clean the texts
similarity_texts <- similarity_corpus %>%
  mutate(text = map(glID, ~gutenberg_download(.x) %>%
  pull(text) %>%
  paste(collapse = " "))) %>%
  mutate(
   text_clean = str_to_lower(text) %>%
      str_remove_all("'s\\b") %>%                    # Remove possessive 's
      str_remove_all("[[:punct:]]") %>%              # Remove punctuation
      str_squish()                                    # Remove extra whitespace
  ) %>%
  select(author, title, year, genre, text_clean)

print(paste("Loaded", nrow(similarity_texts), "texts for similarity analysis"))
```

## Creating TF-IDF Vectors for Similarity

To calculate distances between texts, we need to represent each text as a vector in a common space. We'll create a TF-IDF matrix where:
- Each row represents a text
- Each column represents a unique word in our corpus
- Each cell contains the TF-IDF score for that word in that text

```{r}
# Create TF-IDF matrix for similarity calculations
corpus_words <- similarity_texts %>%
  select(title, author, genre, text_clean) %>%
  unnest_tokens(word, text_clean) %>%
  # Apply lemmatization for better thematic grouping
  mutate(word = lemmatize_words(word)) %>%
  anti_join(stop_words, by = "word") %>%
  # Filter out problematic words that could conflict with column names
  filter(!word %in% c("title", "author", "genre", "text", "word")) %>%
  # Count words and aggregate after lemmatization (some words may become duplicates)
  count(title, word, sort = TRUE) %>%
  # Filter out very rare words (appear in fewer than 2 texts) for cleaner analysis
  group_by(word) %>%
  filter(n_distinct(title) >= 2) %>%
  ungroup()

# Check for any remaining duplicates before creating matrix
duplicate_check <- corpus_words %>%
  count(title, word) %>%
  filter(n > 1)

if(nrow(duplicate_check) > 0) {
  print("Warning: Found duplicate title-word combinations:")
  print(duplicate_check)
}

# Calculate TF-IDF scores
corpus_tfidf <- corpus_words %>%
  bind_tf_idf(word, title, n)

# Create TF-IDF matrix (texts as rows, words as columns)
tfidf_matrix <- corpus_tfidf %>%
  select(title, word, tf_idf) %>%
  # Ensure no duplicates by taking the maximum tf_idf if duplicates exist
  group_by(title, word) %>%
  summarise(tf_idf = max(tf_idf), .groups = "drop") %>%
  pivot_wider(names_from = word, values_from = tf_idf, values_fill = 0) %>%
  column_to_rownames("title") %>%
  as.matrix()

print(paste("Created TF-IDF matrix with", nrow(tfidf_matrix), "texts and", ncol(tfidf_matrix), "words"))
```

## Cosine Distance

**Cosine distance** measures the angle between two vectors, focusing on the *direction* rather than the *magnitude*. It's particularly useful for text analysis because it emphasizes proportional word usage patterns rather than text length.

### Mathematical Foundation

For two vectors **a** and **b**, cosine similarity is:

$$
\text{cosine similarity} = \frac{\mathbf{a} \cdot \mathbf{b}}{|\mathbf{a}| \times |\mathbf{b}|}
$$

And cosine distance is:

$$
\text{cosine distance} = 1 - \text{cosine similarity}
$$

Values range from 0 to 1:
- **0**: Identical direction (highest similarity)
- **1**: Completely opposite direction (lowest similarity)

```{r}
# Calculate cosine distance matrix
cosine_distance <- function(matrix) {
  # Normalize each row (document) by its magnitude
  norms <- sqrt(rowSums(matrix^2))
  # Avoid division by zero
  norms[norms == 0] <- 1
  normalized_matrix <- matrix / norms
  
  # Calculate cosine similarity matrix
  cosine_sim <- normalized_matrix %*% t(normalized_matrix)
  
  # Convert to distance (1 - similarity)
  cosine_dist <- 1 - cosine_sim
  return(cosine_dist)
}

cosine_dist_matrix <- cosine_distance(tfidf_matrix)

# Display the cosine distance matrix
round(cosine_dist_matrix, 3) %>%
  knitr::kable(caption = "Cosine Distance Matrix (0 = identical, 1 = completely different)")
```

Let's visualize the cosine distances as a heatmap to see patterns:

```{r}
#| fig-height: 8
# Create a more readable heatmap for cosine distances
cosine_dist_df <- cosine_dist_matrix %>%
  as.data.frame() %>%
  rownames_to_column("text1") %>%
  pivot_longer(-text1, names_to = "text2", values_to = "cosine_distance") %>%
  # Add genre information
  left_join(similarity_texts %>% select(title, genre) %>% rename(text1 = title, genre1 = genre), by = "text1") %>%
  left_join(similarity_texts %>% select(title, genre) %>% rename(text2 = title, genre2 = genre), by = "text2") %>%
  # Create shorter titles for better visualization
  mutate(
    text1_short = case_when(
      str_detect(text1, "Pride") ~ "Pride & Prejudice",
      str_detect(text1, "Frankenstein") ~ "Frankenstein",
      str_detect(text1, "Alice") ~ "Alice in Wonderland",
      str_detect(text1, "Twenty") ~ "20,000 Leagues",
      str_detect(text1, "Treasure") ~ "Treasure Island",
      str_detect(text1, "Dracula") ~ "Dracula",
      str_detect(text1, "Time Machine") ~ "Time Machine",
      str_detect(text1, "Wind") ~ "Wind in the Willows",
      str_detect(text1, "Gatsby") ~ "Great Gatsby",
      str_detect(text1, "Great Expectations") ~ "Great Expectations",
      TRUE ~ text1
    ),
    text2_short = case_when(
      str_detect(text2, "Pride") ~ "Pride & Prejudice",
      str_detect(text2, "Frankenstein") ~ "Frankenstein",
      str_detect(text2, "Alice") ~ "Alice in Wonderland",
      str_detect(text2, "Twenty") ~ "20,000 Leagues",
      str_detect(text2, "Treasure") ~ "Treasure Island",
      str_detect(text2, "Dracula") ~ "Dracula",
      str_detect(text2, "Time Machine") ~ "Time Machine",
      str_detect(text2, "Wind") ~ "Wind in the Willows",
      str_detect(text2, "Gatsby") ~ "Great Gatsby",
      str_detect(text2, "Great Expectations") ~ "Great Expectations",
      TRUE ~ text2
    )
  )

ggplot(cosine_dist_df, aes(x = text1_short, y = text2_short, fill = cosine_distance)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", 
                       midpoint = 0.5, name = "Cosine\nDistance") +
  labs(
    title = "Cosine Distance Between Texts",
    subtitle = "Blue = More Similar, Red = More Different",
    x = "Text 1", y = "Text 2"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 9),
    axis.text.y = element_text(size = 9),
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5)
  )
```

Now let's find the most and least similar text pairs using cosine distance:

```{r}
# Find most similar and least similar pairs (excluding self-comparisons)
cosine_pairs <- cosine_dist_df %>%
  filter(text1 != text2) %>%
  # Remove duplicate pairs (A-B and B-A are the same)
  filter(text1 < text2) %>%
  arrange(cosine_distance)

# Most similar pairs
cat("MOST SIMILAR TEXTS (Cosine Distance):\n")
head(cosine_pairs, 5) %>%
  select(text1_short, text2_short, cosine_distance, genre1, genre2) %>%
  mutate(cosine_distance = round(cosine_distance, 3)) %>%
  knitr::kable(col.names = c("Text 1", "Text 2", "Cosine Distance", "Genre 1", "Genre 2"))

cat("\nLEAST SIMILAR TEXTS (Cosine Distance):\n")
tail(cosine_pairs, 5) %>%
  select(text1_short, text2_short, cosine_distance, genre1, genre2) %>%
  mutate(cosine_distance = round(cosine_distance, 3)) %>%
  knitr::kable(col.names = c("Text 1", "Text 2", "Cosine Distance", "Genre 1", "Genre 2"))
```

## Euclidean Distance

**Euclidean distance** measures the straight-line distance between two points in space. Unlike cosine distance, it considers both the direction and magnitude of vectors, making it sensitive to the actual values of TF-IDF scores.

### Mathematical Foundation

For two vectors **a** and **b**, Euclidean distance is:

$$
\text{Euclidean distance} = \sqrt{\sum_{i=1}^{n} (a_i - b_i)^2}
$$

Where $n$ is the number of dimensions (words in our vocabulary).

```{r}
# Calculate Euclidean distance matrix
euclidean_dist_matrix <- as.matrix(dist(tfidf_matrix, method = "euclidean"))

# Display part of the Euclidean distance matrix
round(euclidean_dist_matrix, 3) %>%
  knitr::kable(caption = "Euclidean Distance Matrix (smaller = more similar)")
```

Let's visualize the Euclidean distances:

```{r}
#| fig-height: 8
# Create heatmap for Euclidean distances
euclidean_dist_df <- euclidean_dist_matrix %>%
  as.data.frame() %>%
  rownames_to_column("text1") %>%
  pivot_longer(-text1, names_to = "text2", values_to = "euclidean_distance") %>%
  # Add genre information and short titles
  left_join(similarity_texts %>% select(title, genre) %>% rename(text1 = title, genre1 = genre), by = "text1") %>%
  left_join(similarity_texts %>% select(title, genre) %>% rename(text2 = title, genre2 = genre), by = "text2") %>%
  mutate(
    text1_short = case_when(
      str_detect(text1, "Pride") ~ "Pride & Prejudice",
      str_detect(text1, "Frankenstein") ~ "Frankenstein",
      str_detect(text1, "Alice") ~ "Alice in Wonderland",
      str_detect(text1, "Twenty") ~ "20,000 Leagues",
      str_detect(text1, "Treasure") ~ "Treasure Island",
      str_detect(text1, "Dracula") ~ "Dracula",
      str_detect(text1, "Time Machine") ~ "Time Machine",
      str_detect(text1, "Wind") ~ "Wind in the Willows",
      str_detect(text1, "Gatsby") ~ "Great Gatsby",
      str_detect(text1, "Great Expectations") ~ "Great Expectations",
      TRUE ~ text1
    ),
    text2_short = case_when(
      str_detect(text2, "Pride") ~ "Pride & Prejudice",
      str_detect(text2, "Frankenstein") ~ "Frankenstein",
      str_detect(text2, "Alice") ~ "Alice in Wonderland",
      str_detect(text2, "Twenty") ~ "20,000 Leagues",
      str_detect(text2, "Treasure") ~ "Treasure Island",
      str_detect(text2, "Dracula") ~ "Dracula",
      str_detect(text2, "Time Machine") ~ "Time Machine",
      str_detect(text2, "Wind") ~ "Wind in the Willows",
      str_detect(text2, "Gatsby") ~ "Great Gatsby",
      str_detect(text2, "Great Expectations") ~ "Great Expectations",
      TRUE ~ text2
    )
  )

ggplot(euclidean_dist_df, aes(x = text1_short, y = text2_short, fill = euclidean_distance)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", 
                       midpoint = median(euclidean_dist_df$euclidean_distance), 
                       name = "Euclidean\nDistance") +
  labs(
    title = "Euclidean Distance Between Texts",
    subtitle = "Blue = More Similar, Red = More Different",
    x = "Text 1", y = "Text 2"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 9),
    axis.text.y = element_text(size = 9),
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5)
  )
```

Let's find the most and least similar pairs using Euclidean distance:

```{r}
# Find most similar and least similar pairs using Euclidean distance
euclidean_pairs <- euclidean_dist_df %>%
  filter(text1 != text2) %>%
  filter(text1 < text2) %>%
  arrange(euclidean_distance)

# Most similar pairs
cat("MOST SIMILAR TEXTS (Euclidean Distance):\n")
head(euclidean_pairs, 5) %>%
  select(text1_short, text2_short, euclidean_distance, genre1, genre2) %>%
  mutate(euclidean_distance = round(euclidean_distance, 3)) %>%
  knitr::kable(col.names = c("Text 1", "Text 2", "Euclidean Distance", "Genre 1", "Genre 2"))

cat("\nLEAST SIMILAR TEXTS (Euclidean Distance):\n")
tail(euclidean_pairs, 5) %>%
  select(text1_short, text2_short, euclidean_distance, genre1, genre2) %>%
  mutate(euclidean_distance = round(euclidean_distance, 3)) %>%
  knitr::kable(col.names = c("Text 1", "Text 2", "Euclidean Distance", "Genre 1", "Genre 2"))
```

## Comparing Cosine vs. Euclidean Distance

Let's directly compare how these two distance measures rank text similarities:

```{r}
# Compare the two distance measures
distance_comparison <- cosine_pairs %>%
  select(text1, text2, cosine_distance) %>%
  left_join(
    euclidean_pairs %>% select(text1, text2, euclidean_distance),
    by = c("text1", "text2")
  ) %>%
  # Calculate rankings for each distance measure
  mutate(
    cosine_rank = rank(cosine_distance),
    euclidean_rank = rank(euclidean_distance),
    rank_difference = abs(cosine_rank - euclidean_rank)
  ) %>%
  left_join(similarity_texts %>% select(title, genre) %>% rename(text1 = title, genre1 = genre), by = "text1") %>%
  left_join(similarity_texts %>% select(title, genre) %>% rename(text2 = title, genre2 = genre), by = "text2")

# Show pairs where the rankings differ most
cat("PAIRS WITH BIGGEST RANKING DIFFERENCES:\n")
distance_comparison %>%
  arrange(desc(rank_difference)) %>%
  head(10) %>%
  mutate(
    text1_short = str_trunc(text1, 20),
    text2_short = str_trunc(text2, 20)
  ) %>%
  select(text1_short, text2_short, cosine_rank, euclidean_rank, rank_difference, genre1, genre2) %>%
  knitr::kable(col.names = c("Text 1", "Text 2", "Cosine Rank", "Euclidean Rank", "Difference", "Genre 1", "Genre 2"))
```

Let's create a scatter plot to visualize the relationship between the two distance measures:

```{r}
#| fig-height: 8
# Scatter plot comparing the two distance measures
distance_comparison %>%
  mutate(
    same_genre = genre1 == genre2,
    pair_label = ifelse(rank_difference >= 10, 
                       paste(str_trunc(text1, 15), "vs", str_trunc(text2, 15)), "")
  ) %>%
  ggplot(aes(x = cosine_distance, y = euclidean_distance, color = same_genre)) +
  geom_point(size = 3, alpha = 0.7) +
  geom_text(aes(label = pair_label), hjust = 0, vjust = 0, size = 3, check_overlap = TRUE) +
  geom_smooth(method = "lm", se = FALSE, color = "black", linetype = "dashed") +
  labs(
    title = "Cosine vs. Euclidean Distance: How Do They Compare?",
    subtitle = "Each point represents a pair of texts",
    x = "Cosine Distance",
    y = "Euclidean Distance",
    color = "Same Genre"
  ) +
  theme_minimal() +
  scale_color_manual(values = c("FALSE" = "red", "TRUE" = "blue"))
```

## Hierarchical Clustering with Distance Measures

One powerful application of distance measures is **hierarchical clustering** - automatically grouping texts based on their similarities. Let's create dendrograms using both distance measures:

```{r}
#| fig-height: 10
# Create hierarchical clustering dendrograms
par(mfrow = c(2, 1))

# Cosine distance clustering
cosine_hclust <- hclust(as.dist(cosine_dist_matrix), method = "ward.D2")
plot(cosine_hclust, main = "Hierarchical Clustering - Cosine Distance", 
     xlab = "Texts", ylab = "Distance", cex = 0.8)

# Euclidean distance clustering  
euclidean_hclust <- hclust(as.dist(euclidean_dist_matrix), method = "ward.D2")
plot(euclidean_hclust, main = "Hierarchical Clustering - Euclidean Distance", 
     xlab = "Texts", ylab = "Distance", cex = 0.8)

par(mfrow = c(1, 1))
```

Let's examine how well these clusters align with our known genres:

```{r}
# Create clusters and compare with actual genres
# Cut the dendrograms to create groups (6 genres in our focused subset)
cosine_clusters <- cutree(cosine_hclust, k = 6)  
euclidean_clusters <- cutree(euclidean_hclust, k = 6)

# Create comparison table
cluster_comparison <- data.frame(
  Title = names(cosine_clusters),
  Actual_Genre = similarity_texts$genre[match(names(cosine_clusters), similarity_texts$title)],
  Cosine_Cluster = cosine_clusters,
  Euclidean_Cluster = euclidean_clusters
) %>%
  mutate(
    Title_Short = case_when(
      str_detect(Title, "Pride") ~ "Pride & Prejudice",
      str_detect(Title, "Frankenstein") ~ "Frankenstein",
      str_detect(Title, "Alice") ~ "Alice in Wonderland",
      str_detect(Title, "Twenty") ~ "20,000 Leagues",
      str_detect(Title, "Treasure") ~ "Treasure Island",
      str_detect(Title, "Dracula") ~ "Dracula",
      str_detect(Title, "Time Machine") ~ "Time Machine",
      str_detect(Title, "Wind") ~ "Wind in the Willows",
      str_detect(Title, "Gatsby") ~ "Great Gatsby",
      str_detect(Title, "Great Expectations") ~ "Great Expectations",
      TRUE ~ Title
    )
  ) %>%
  arrange(Actual_Genre, Title_Short)

knitr::kable(cluster_comparison %>% select(-Title), 
             col.names = c("Book", "Actual Genre", "Cosine Cluster", "Euclidean Cluster"),
             caption = "Clustering Results vs. Actual Genres")
```

## Genre Similarity Analysis

Finally, let's examine how our distance measures capture similarities within and between literary genres:

```{r}
#| fig-height: 8
# Calculate average distances within and between genres
genre_distances <- cosine_dist_df %>%
  filter(text1 != text2) %>%
  # Add euclidean distance data
  left_join(euclidean_dist_df %>% select(text1, text2, euclidean_distance), 
            by = c("text1", "text2")) %>%
  mutate(
    relationship = case_when(
      genre1 == genre2 ~ paste("Within", genre1),
      TRUE ~ paste(pmin(genre1, genre2), "vs", pmax(genre1, genre2))
    )
  ) %>%
  group_by(relationship) %>%
  summarise(
    avg_cosine_distance = mean(cosine_distance),
    avg_euclidean_distance = mean(euclidean_distance, na.rm = TRUE),
    n_pairs = n(),
    .groups = "drop"
  ) %>%
  arrange(avg_cosine_distance)

# Visualize genre relationships
genre_distances %>%
  pivot_longer(cols = c(avg_cosine_distance, avg_euclidean_distance), 
               names_to = "distance_type", values_to = "distance") %>%
  mutate(
    distance_type = case_when(
      distance_type == "avg_cosine_distance" ~ "Cosine Distance",
      distance_type == "avg_euclidean_distance" ~ "Euclidean Distance"
    ),
    within_genre = str_detect(relationship, "Within")
  ) %>%
  ggplot(aes(x = reorder(relationship, distance), y = distance, 
             fill = within_genre, alpha = distance_type)) +
  geom_col(position = "dodge") +
  scale_alpha_manual(values = c("Cosine Distance" = 0.8, "Euclidean Distance" = 0.6)) +
  scale_fill_manual(values = c("TRUE" = "darkgreen", "FALSE" = "darkred")) +
  coord_flip() +
  labs(
    title = "Average Distances Within and Between Genres",
    subtitle = "Do texts cluster by literary genre?",
    x = "Genre Relationship",
    y = "Average Distance",
    fill = "Within Genre",
    alpha = "Distance Type"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## Summary and Interpretation

```{r}
# Summary statistics
cat("SUMMARY OF FINDINGS:\n")
cat("====================\n\n")

# Overall correlation between distance measures
correlation <- cor(distance_comparison$cosine_distance, distance_comparison$euclidean_distance)
cat("Correlation between Cosine and Euclidean distances:", round(correlation, 3), "\n\n")

# Within-genre vs between-genre similarities
within_genre_avg_cosine <- mean(genre_distances$avg_cosine_distance[str_detect(genre_distances$relationship, "Within")])
between_genre_avg_cosine <- mean(genre_distances$avg_cosine_distance[!str_detect(genre_distances$relationship, "Within")])

cat("Average Cosine Distance:\n")
cat("  Within genres:", round(within_genre_avg_cosine, 3), "\n")
cat("  Between genres:", round(between_genre_avg_cosine, 3), "\n")
cat("  Difference:", round(between_genre_avg_cosine - within_genre_avg_cosine, 3), "\n\n")

# Most distinctive genres (smallest within-genre distances)
most_cohesive_genre <- genre_distances %>%
  filter(str_detect(relationship, "Within")) %>%
  slice_min(avg_cosine_distance, n = 1)

cat("Most cohesive genre (by cosine distance):", most_cohesive_genre$relationship, 
    "with average distance", round(most_cohesive_genre$avg_cosine_distance, 3), "\n")
```

::: {.callout-note title="Key Insights from Distance Analysis"}

Our exploration of text similarity reveals several important patterns:

**1. Distance Measure Differences:**
- **Cosine distance** focuses on proportional word usage patterns, making it excellent for identifying thematic similarities regardless of text length
- **Euclidean distance** considers both proportions and magnitudes, making it sensitive to vocabulary intensity and text scale

**2. Genre Clustering:**
- Texts within the same genre generally show smaller distances (higher similarity) than texts from different genres
- However, some cross-genre similarities emerge, suggesting shared narrative techniques or thematic elements

**3. Mathematical vs. Literary Categories:**
- While distance measures often align with literary genres, they sometimes reveal unexpected similarities based on vocabulary patterns
- This demonstrates how computational methods can both confirm and challenge traditional literary classifications

**4. Applications:**
- **Authorship attribution**: Comparing unknown texts to known authors
- **Genre classification**: Automatically categorizing texts by similarity to genre exemplars  
- **Literary influence**: Detecting vocabulary patterns that suggest textual relationships
- **Corpus exploration**: Discovering unexpected connections in large text collections
:::

These similarity measures provide the foundation for more advanced techniques like clustering, topic modeling, and machine learning applications in literary analysis. In our next episode, we'll explore how these concepts extend into **topic modeling** and **latent semantic analysis**.
